---
title: "Multivarite Normal Distribution"
author: "Daniel Emaasit"
date: "March 17, 2015"
output: html_document
---

# Problem 4.27:
Consider the radiation data in Example 4.10.  

Construct a Q-Q plot for the natural logarithms of these data. Do the natural logarithms appear to normally distributed? 

```{r}
# create a data matrix
radiation <- matrix(c(0.15,0.9,0.18,0.1,0.05,0.12,0.08,0.05,0.08,0.1,0.07,0.02,0.01,0.1,0.1,0.1,0.02,0.1,0.01,0.4,0.1,0.05,0.03,0.05,0.15,0.1,0.15,0.09,0.08,0.18,0.1,0.2,0.11,0.3,0.02,0.2,0.2,0.3,0.3,0.4,0.3,0.05))

colnames(radiation) <-c("Radiation")

# construct a q-q plot
qqnorm(log(radiation))
qqline(log(radiation))

# compute the marginal normality
shapiro.test(log(radiation))
```

The Q-Q plot appears to be straight, hence the data are normally distributed. In addition, the test of normality produces a p-value of 0.1665 showing that the data is normally distributed.  

For lambda = 1/4, the coefficient of normality is greater than for lambda = 0.  

# Problem 4.29
Given the air pollution data.

```{r}
# read the data into R
pollution_data <- read.table("table15.txt")

# examine columns five and six
data2 <- (pollution_data[,5:6])
colnames(data2) <- c("NO2", "O3")

# calcualate the sample mean vector
cm <- colMeans(data2)
cm

# calculate the variance-covariance Matrix
S <- cov(data2)
S

# calaculate the statistical square distances
d <- apply(data2, MARGIN = 1, function(data2) +
                   t(data2-cm) %*% solve(S) %*% (data2-cm))
d

# construct a chi-square plot of the ordered distances
par(mfrow = c(1,1))
plot(qchisq((1:nrow(data2)-1/2)/nrow(data2),df=2), sort(d),
     xlab = expression(paste(chi[2]^2, "Quantile")),
     ylab = "Ordered Distances",
     main = "Chi-square Plot for (NO2, O3)"); abline(a=0, b=1)
```

The number of units = 26 
The proportion = 62%

# Problem 4.33

## (a)
Examine the margninal normality of the observations for the data in Table 4.3.

```{r}
# read the data into R
data3 <- read.table("table43.txt")
colnames(data3)<-c("X1", "X2", "X3", "X4","X5")
library(MVN)
uniNorm(data3, type = "SW", desc = TRUE)

```


## (b)
Examine the bivariate normality of the observations for the data.

```{r}
# examine columns one upto four
data3 <- (data3[,-5])

# construct the bivariate chi-square plot for X1 and X2
data3_X1x2 <- (data3[,c("X1","X2")])
cm3_X1X2 <- colMeans(data3_X1x2)
S3_X1X2 <- cov(data3_X1x2)
d3_X1X2 <- apply(data3_X1x2, MARGIN = 1, function(data3_X1x2) +
                   t(data3_X1x2-cm3_X1X2) %*% solve(S3_X1X2) %*% (data3_X1x2-cm3_X1X2))

par(mfrow = c(3,2))
plot(qchisq((1:nrow(data3_X1x2)-1/2)/nrow(data3_X1x2),df=2), sort(d3_X1X2),
     xlab = expression(paste(chi[2]^2, "Quantile")),
     ylab = "Ordered Distances",
     main = "Chi-Square Plot for (X1,X2)"); abline(a=0, b=1)

# construct the bivariate chi-square plot for X1 and X3
data3_X1x3 <- (data3[,c("X1","X3")])
cm3_X1X3 <- colMeans(data3_X1x3)
S3_X1X3 <- cov(data3_X1x3)
d3_X1X3 <- apply(data3_X1x3, MARGIN = 1, function(data3_X1x3) +
                   t(data3_X1x3-cm3_X1X3) %*% solve(S3_X1X3) %*% (data3_X1x3-cm3_X1X3))

plot(qchisq((1:nrow(data3_X1x3)-1/2)/nrow(data3_X1x3),df=2), sort(d3_X1X3),
     xlab = expression(paste(chi[2]^2, "Quantile")),
     ylab = "Ordered Distances",
     main = "Chi-Square Plot for (X1,X3)"); abline(a=0, b=1)

# construct the bivariate chi-square plot for X1 and X4
data3_X1x4 <- (data3[,c("X1","X4")])
cm3_X1X4 <- colMeans(data3_X1x4)
S3_X1X4 <- cov(data3_X1x4)
d3_X1X4 <- apply(data3_X1x4, MARGIN = 1, function(data3_X1x4) +
                   t(data3_X1x4-cm3_X1X4) %*% solve(S3_X1X4) %*% (data3_X1x4-cm3_X1X4))

plot(qchisq((1:nrow(data3_X1x4)-1/2)/nrow(data3_X1x4),df=2), sort(d3_X1X4),
     xlab = expression(paste(chi[2]^2, "Quantile")),
     ylab = "Ordered Distances",
     main = "Chi-Square Plot for (X1,X4)"); abline(a=0, b=1)

# construct the bivariate chi-square plot for X2 and X3
data3_X2x3 <- (data3[,c("X2","X3")])
cm3_X2x3 <- colMeans(data3_X2x3)
S3_X2x3 <- cov(data3_X2x3)
d3_X2x3 <- apply(data3_X2x3, MARGIN = 1, function(data3_X2x3) +
                   t(data3_X2x3-cm3_X2x3) %*% solve(S3_X2x3) %*% (data3_X2x3-cm3_X2x3))


plot(qchisq((1:nrow(data3_X2x3)-1/2)/nrow(data3_X2x3),df=2), sort(d3_X2x3),
     xlab = expression(paste(chi[2]^2, "Quantile")),
     ylab = "Ordered Distances",
     main = "Chi-Square Plot for (X2,X3)"); abline(a=0, b=1)

# construct the bivariate chi-square plot for X2 and X4
data3_X2x4 <- (data3[,c("X2","X4")])
cm3_X2x4 <- colMeans(data3_X2x4)
S3_X2x4 <- cov(data3_X2x4)
d3_X2x4 <- apply(data3_X2x4, MARGIN = 1, function(data3_X2x4) +
                   t(data3_X2x4-cm3_X2x4) %*% solve(S3_X2x4) %*% (data3_X2x4-cm3_X2x4))


plot(qchisq((1:nrow(data3_X2x4)-1/2)/nrow(data3_X2x4),df=2), sort(d3_X2x4),
     xlab = expression(paste(chi[2]^2, "Quantile")),
     ylab = "Ordered Distances",
     main = "Chi-Square Plot for (X2,X4)"); abline(a=0, b=1)

# construct the bivariate chi-square plot for X3 and X4
data3_X3x4 <- (data3[,c("X3","X4")])
cm3_X3x4 <- colMeans(data3_X3x4)
S3_X3x4 <- cov(data3_X3x4)
d3_X3x4 <- apply(data3_X3x4, MARGIN = 1, function(data3_X3x4) +
                   t(data3_X3x4-cm3_X3x4) %*% solve(S3_X3x4) %*% (data3_X3x4-cm3_X3x4))


plot(qchisq((1:nrow(data3_X3x4)-1/2)/nrow(data3_X3x4),df=2), sort(d3_X3x4),
     xlab = expression(paste(chi[2]^2, "Quantile")),
     ylab = "Ordered Distances",
     main = "Chi-Square Plot for (X3,X4)"); abline(a=0, b=1)
```

The chi-square plots for (x1,x2), (x1,x3), and (x3,x4) are straight lines while (x1,x4), (x2,x3) and (x2,x4) do not appear to be straight.

# Problem 4.39

## (a)
Using the data in Table 4.6, examine each of the variables independence, support, benevolence, conformity, and leadership for marginal normality.

```{r}
# read the data into R
data4 <- read.table("table46.txt")

# examine columns one upto five
data4 <- (data4[,1:5])
colnames(data4) <- c("independence", "support", "benevolence", "conformity", "leadership")

# test the marginal normality for each variable
uniNorm(data4, type = "SW", desc = TRUE)
```

Only two variables are normaly distributed, that is, benevolence and conformity.  

## (b)
Using all five variables, check for multivariate normality

```{r}
# calculate the sample mean vector
cm4 <- colMeans(data4)

# calculate the variance-covariance Matrix
S4 <- cov(data4)

# calaculate the statistical square distances
d4 <- apply(data4, MARGIN = 1, function(data4) +
                   t(data4-cm4) %*% solve(S4) %*% (data4-cm4))

# construct the chi-square plots
plot(qchisq((1:nrow(data4)-1/2)/nrow(data4),df=5), sort(d4),
     xlab = expression(paste(chi[5]^2, "Quantile")),
     ylab = "Ordered Distances",
     main = "The Chi-Squared Plot for all Variables in Psychological Data"); abline(a=0, b=1)
```

The chi-square plot is not straight thereby showing non-normality for all five variables  

## (c)
For those variables that are not normal, determine the transformation that makes them nearly normal  

```{r}
# take the square root of the variables independence and leadership
data4$independence <- sqrt(data4$independence)
data4$leadership <- sqrt(data4$leadership)

# take the natural log of variable support
data4$support <- log(data4$support)

# test for normality
uniNorm(data4, type = "SW", desc = TRUE)
```

Now all the variables are normaly distributed except for support.

# Problem 4.41

Consider the data on snow removal.

## (a)
Comment on any outliers in a scatter plot of the original variables

```{r}
# read the data into R
data5 <- read.table("table32.txt")
colnames(data5) <- c("Duration", "Machine_Time")

# scatter plot of the data
library(ggplot2)
ggplot(data5, aes(Machine_Time, Duration)) + geom_point(color = "blue", size = 4) + ggtitle ("Scatter Plot of Original Variables in the Snow Data")
```

There appears to be one outlier at position (42.3, 17.5) 

## (b)
Determine the power transformation that makes "Duration" approximately normal. Construct a QQ plot of the transformed observations

```{r}
# first check for normality before transformation
shapiro.test(data5$Duration)

# transform by taking the natural log of Duration i.e (lambda = 0)
shapiro.test(log(data5$Duration))

# construct a QQ plot of the transformed Duration observations 
qqnorm(log(data5$Duration))
qqline(log(data5$Duration))
```


## (c)
Determine the power transformation that makes "Machine_Time" approximately normal. Construct a QQ plot of the transformed observations

```{r}
# first check for normality before transformation
shapiro.test(data5$Machine_Time)

# transform by taking the natural log of Machine_Time data i.e (lambda = 0)
shapiro.test(log(data5$Machine_Time))

# construct a QQ plot of the transformed Machine_Time data
qqnorm(log(data5$Machine_Time))
qqline(log(data5$Machine_Time))
```

## (d)
Determine the power transformation for approximate bivariate normality using (4-40)

```{r}

# calculate the sample mean vector
data5 <- log(data5)
cm5 <- colMeans(data5)

# calculate the variance-covariance Matrix
S5 <- cov(data5)

# calaculate the statistical square distances
d5 <- apply(data5, MARGIN = 1, function(data5) +
                   t(data5-cm5) %*% solve(S5) %*% (data5-cm5))

# construct the chi-square plots
plot(qchisq((1:nrow(data5)-1/2)/nrow(data5),df=2), sort(d5),
     xlab = expression(paste(chi[2]^2, "Quantile")),
     ylab = "Ordered Distances",
     main = "The Chi-Squared Plot for Transformed Snow Data"); abline(a=0, b=1)
```
